Using best practices for forecasting biodiversity changes in breeding birds

### 2017-08-06 manuscript/draft.Rmd
barcharts etc

We fit three sets of models that were fit to each site independently, with no
environmental information. These models were fit to each BBS route both on the
raw richness values and on the residuals from our observer model. When
correcting for observer effects, we averaged across 500 models that were fit
separately to the 500 Monte Carlo estimates of the observer effects, to account
for our uncertainty in the true values of those effects. All of these models use
a Gaussian error distribution.

@@

We evaluated each model's for forecasts using the data for each year between `r
timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
on average, the models' predictions were from the observed value; 2) the 95%
prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit incorporating good point
estimates, precision, and coverage.  In addition to looking at forecast
performance in general we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against the forecast
horizon.

@@

In general, each model's error increased more-or-less linearly with time,
although the slopes and initial error rates varied across models (Figure 5). The
mistnet JSDM and GBM-based macroecological model's deviance tended to rise less
rapidly over time than the baseline models' deviance did; for the ten-year
forecasts, these environment-based models outperformed the Naive model and
nearly matched the performance of the Average model.

When the differences among observers were not explicitly accounted for, nearly
all metrics of fit worsened (Figure 6). The increased error resulted from a
small number of forecasts where observer turnover caused a large shift in the
reported richness values. The Naive model was less sensitive to these shifts,
because it largely ignored the richness values reported by observers that had
retired by the end of the training period (Figure 1). The Average model, which
gave equal weight to observations from the whole training period, showed a
larger decline in performance -- especially in terms of coverage. The mistnet
JSDM was notable here because the coverage of its prediction intervals retained
very good coverage in the presence of this observer-based noise, which we
attribute to its ability to model this variation with its latent variables.

![Accounting for differences between observers typically has a small impact on
accuracy (especially for the "naive" model), but occasonally changes the
predicted richness values by more than 10 species. These changes usually improve
the model's predictions, but not always.](../figures/observers.png)

@@

The most commonly used method to forecast future biodiversity, stacked SDMs
[@urban2015], provided the worst overall forecasts of all methods evaluated. In
general, SDM-based forecasts are rarely evaluated, particularly in terms of
repeated observations at the same site. Since site-level richness is relatively
stable, model evaluations that show high accuracy across spatial gradients may
not indicate much about a model's ability to predict changes over time. This
issue has also recently been highlighted as an issue for using SDMs to predict
the future locations of individual species [@rapacciuolo2012]. This result is
particularly sobering because this approach forms the main foundation for
estimates of the predicted loss of biodiversity to climate change [@thomas2004;
@urban2015]. The poor performance of SDM-based forecasts highlights the crucial
importance of comparing multiple modeling approaches when conducting ecological
forecasts, and in particular the value of comparing results to simple baselines
to avoid over-interpreting the information present in these forecasts (Box 1).
Disciplines that have more mature forecasting cultures often do this by
reporting "forecast skill", i.e., the improvement in the forecast relative to a
simple baseline [@jolliffe2003]. We recommend following Ye et al. [@ye2015] and
adopting this approach in future ecological forecasting research. It also
highlights the need for using hindcasting and time-series data for assessing the
effectiveness of forecasting methods, since this allows comparisons to baselines
and assessment of predicted dynamics within sites or regions.

@@

For the error metric that evaluates both models' point estimates and their
uncertainty simultaneously, the JSDM and the nonlinear richness model converge
towards the time-series baselines at the longest time-scales assessed. This
convergence at longer time scales highlights the importance of considering how
forecasting method performance changes with the distance into the future being
predicted (Box 1). The value of this approach, and associated forecast horizons,
has been raised for knowing how far into the future a model can be effectively
used for forecasting [@petchey2015], but our results suggest a broader value to
this approach for considering the potential importance of different models and
processes for making forecasts at different scales. Our results show that for
forecasts 1-3 years in the future, baselines and time-series based approaches
outperform all other methods using integrative metrics. At these time scales,
environmental changes are relatively small and changes in the biota may lag
behind changes in the environment. However, as the timescale of the forecast
increases to a decade the JSDMs and environmental richness models get closer
to the accuracy of the time-series methods and even surpass some of those
methods. As the forecast distance increases the amount of environmental change
is expected to increase and the system will have more time to respond,
potentially leading to shifts towards improved relative performance of the
models incorporating environmental and ecological information. If this shift
with forecast distance in the relative importance of different processes and
models continues, this could lead to JSDMs and environmental richness models
outperforming time-series based approaches at sufficiently long time scales.
These results suggest that assessment of how forecast performance changes with
time lag is not only important for determining how far in the future to
forecast, but also for determining which models and processes are most relevant
for making forecasts at different time scales. For example ensemble forecasts
could be created which incorporate these time lags into the weights for
individual models.

>>>>

We fit three sets of models that were fit to each site independently, with no
environmental information. These models were fit to each BBS route both on the
raw richness values and on the residuals from our observer model. When
correcting for observer effects, we averaged across 500 models that were fit
separately to the 500 Monte Carlo estimates of the observer effects, to account
for our uncertainty in the true values of those effects. All of these models use
a Gaussian error distribution (rather than a count distribution) for the reasons
discussed below .

@@

We evaluated each model's for forecasts using the data for each year between `r
timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
on average, the models' predictions were from the observed value; 2) the 95%
prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit incorporating good point
estimates, precision, and coverage.  In addition to looking at forecast
performance in general we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against the forecast
horizon. Finally, we decomposed each model's squared error into two components:
one described the squared error associated with site-level means, while the
other described the squared error associated with annual fluctuations in
richness within a site. This decomposition tells us the extent to which each
model's error depends on consistent differences among sites versus changes in
site-level richness from year to year.

@@

Figure 6 shows each model's squared error, divided into consistent site-level
biases, versus errors that fluctuated year-to-year. Across all models, most of
the error came from errors in the site-level mean. The "average" model, which
was based on the site-level mean, had the lowest error in this regard. In
contrast, the three environmental models showed the largest biases at the site
level; this makes sense, given that they could not explicitly distinguish among
sites with similar climate, NDVI, and elevation.

![Each model's squared error can be divided into two parts. If a model
consistently over-estimates or under-estimates the richness at a given site,
this contributes to its error in predicting the site-level mean. The second
component is based on errors in predicting fluctuations in a site's richness
over time. Both components of the mean squared error were lower for the time
series models than for the environmental models.](../figures/barcharts.png)

When the differences among observers were not explicitly accounted for, nearly
all metrics of fit worsened (Figure 7). The increased error resulted from a
small number of forecasts where observer turnover caused a large shift in the
reported richness values. The Naive model was less sensitive to these shifts,
because it largely ignored the richness values reported by observers that had
retired by the end of the training period (Figure 1). The Average model, which
gave equal weight to observations from the whole training period, showed a
larger decline in performance -- especially in terms of coverage. The mistnet
JSDM was notable here because the coverage of its prediction intervals retained
very good coverage in the presence of this observer-based noise, which we
attribute to its ability to model this variation with its latent variables.

![Accounting for differences between observers typically has a small impact on
accuracy (especially for the "naive" model), but occasonally changes a
prediction's accuracy by more than 10 species. These changes are beneficial, on
average.](../figures/observers.png)

@@

The most commonly used method to forecast future biodiversity, SDMs
[@urban2015], provided the worst overall forecasts of all methods evaluated. In
general, SDM-based forecasts are rarely evaluated, particularly in terms of
repeated observations at the same site. Since site-level richness is relatively
stable, model evaluations that show high accuracy across spatial gradients may
not indicate much about a model's ability to predict changes over time. This
issue has also recently been highlighted as an issue for using SDMs to predict
the future locations of individual species [@rapacciuolo2012]. This result is
particularly sobering because this approach forms the main foundation for
estimates of the predicted loss of biodiversity to climate change [@thomas2004;
@urban2015]. The poor performance of SDM-based forecasts highlights the crucial
importance of comparing multiple modeling approaches when conducting ecological
forecasts, and in particular the value of comparing results to simple baselines
to avoid over-interpreting the information present in these forecasts (Box 1).
Disciplines that have more mature forecasting cultures often do this by
reporting "forecast skill", i.e., the improvement in the forecast relative to a
simple baseline [@jolliffe2003]. We recommend following Ye et al. [@ye2015] and
adopting this approach in future ecological forecasting research. It also
highlights the need for using hindcasting and time-series data for assessing the
effectiveness of forecasting methods, since this allows comparisons to baselines
and assessment of predicted dynamics within sites or regions.

@@
